---
title: "Comparison of the algorithms"
author: "Deepak Tanwar & Hana Parizkova"
date: "<b>Created on:</b> 2019-01-07 <br> <b>Updated on:</b> `r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    highlight: pyg
    keep_md: no
    number_sections: yes
    fig_width: 8.5
    fig_height: 11
    fig_caption: true
    df_print: paged
    code_folding: show
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  fontsize: 12pt
  geometry: margin=1in
  documentclass: article
---

# Packages required
```{r, warning=F, message=FALSE}
library(GenomicRanges)
library(ggplot2)
library(data.table)
library(plotly)
```

# Input files

## 100 simulated DMRs
```{r}
read.dmr <- GRanges(read.delim("./input/simulated_DMRS.txt.gz", sep = "\t", stringsAsFactors = F, header = T))
```


## Detected DMRs
Read the DMRs detected in simulated data by the individual methods.
```{r}
files.sim <- list.files(path = "./input", pattern = "sim_", full.names = T)
names(files.sim) <- sapply(files.sim, function(x) strsplit(x, "\\.|/")[[1]][4])

read.sim <- lapply(files.sim, function(x) GRanges(read.delim(x, sep = "\t", stringsAsFactors = F, header = T)))
```

Separately load the data of the two methods that do not provide p-value or q-value estimate (BSseq, MethPipe).
```{r}
files.sim.no.qval <- list.files(path = "./input/input_no_qval", pattern = "sim_", full.names = T)
names(files.sim.no.qval) <- sapply(files.sim.no.qval, function(x) strsplit(x, "\\.|/")[[1]][5])

read.sim.no.qval <- lapply(files.sim.no.qval, function(x) GRanges(read.delim(x, sep = "\t", stringsAsFactors = F, header = T)))
```

Modify the names of objects in `read.sim` and `read.sim.no.qval` to be nicer:
```{r}
modify_name <- function(name){
  if (startsWith(name, "DMRcaller")) {
    name.split <- strsplit(name, "_")[[1]]
    new.name <- paste(name.split[1], head(tail(name.split, n=3), n=1), sep="_")
  } else {
    new.name <- strsplit(name, "_")[[1]][1]
  }
  return(new.name)
}

names(read.sim) <- sapply(names(read.sim), function(x) modify_name(x))
names(read.sim.no.qval) <- sapply(names(read.sim.no.qval), function(x) modify_name(x))
```

# Filtering results for different q-value cut-offs
For several q-value cut-offs we filter out only those regions passing the cut-off.
```{r}
cutoffs <- c(0.001, 0.005, 0.01, 0.05, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9)
sim.new <- list()

# for each cut-off
for (j in 1:length(cutoffs)) {
  sim.fdr <- list()
  # create a list of GRanges objects, one GRanges object for each method 
  for (i in 1:length(read.sim)) {
    n <- names(read.sim)[i]
    sim.fdr[[i]] <- read.sim[[i]][read.sim[[i]]$qval <= cutoffs[[j]]]
    names(sim.fdr)[i] <- n
    colnames(mcols(sim.fdr[[i]])) <- paste(n, colnames(mcols(sim.fdr[[i]])), sep = ":")
  }
  sim.new[[j]] <- sim.fdr
}
```


# Computing the comparison statistics
## Negative control

## Simulated data
### Helper function
Helper function which returns length of overlap of one range (`gr1`) with any of ranges in list `gr2`.

```{r}
#   gr1 ... one range
#   gr2 ... list of ranges
meta_GR_overlap <- function(gr1, gr2) {
  ranges <- overlapsRanges(gr1, gr2, minoverlap = 1)
  return(sum(width(ranges)))
}
```

### Ways of computing TPs and FPs
Three different ways of computing true positive, false positive and false negative DMRs were used:
1. **Any overlap:** A detected DMR is considered to be true positive ($TP_d$) if it has any overlap with any simulated DMR. A detected DMR is considered to be false positive ($FP_d$) if it has no overlap with any simulated DMR. A simulated DMR is considered correctly identified ($TP_s$) if it has any overlap with any detected DMR.

2. **Threshold:** A detected DMR is classified as $TP_d$ if it is overlapped by at least $x$ % of its length by one or more simulated regions, $x$ is a selected threshold (80 % in our case). Otherwise it is classified as $FP_d$. Similarly, a simulated DMR is classified as $TP_s$ it is covered by at least $x$ % by one or more detected DMRs.

3. **Proportional:** If a detected DMR is covered by $x*100$ % of its length by a simulated DMR (and by $(1-x)*100 $ % it is not covered by any simulated DMR), then it adds $x$ to $TP_d$ and $1-x$ to $FP_d$. Similarly, if a simulated DMR is detected from $x*100$ %, then it adds $x$ to $TP_s$.

Observed false discovery rate (FDR) is then computed as $FDR = FP_d / (FP_d + TP_d)$, power is computed as $p = TP_s / 100$ (because there is 100 simulated DMRs in total).

For comparison and for verifying the results (see below) we have computed the FDR and power also based on individual nucleotides (being aware that this is not really correct for DMRs):

4. **Nucleotidewise:** Every nucleotide position (for simplicity we have used all positions, not just C's) is classified as:
  - TP if it is inside a simulated DMR and inside a detected DMR
  - FP if it is not inside a simulated DMR and is inside a detected DMR
  - TN if it is neither inside a simulated nor a detected DMR
  - FN if it is inside a simulated DMR but not inside a detected DMR
  
Observed FDR is then computed as $FDR = FP / (FP + TP)$, power is computed as $p = TP/(TP+FN)$.


### Computation

Threshold for required overlap of regions in the `threshold` method:
```{r}
threshold <- 0.8
```

```{r}
stats.each.threshold <- list()
stats.each.all <- list()
stats.each.prop <- list()
stats.each.nuc <- list()

# total length of all simulated DMRs
tot.length.sim <- sum(width(read.dmr))
```

For every method which provides q-value estimate we compute number of TPs and FPs, FDR, and power as described above.
```{r}
# for every method
for (j in 1:length(sim.new[[1]])) {
  n <- names(sim.new[[1]])[j]
  data.tmp.threshold <- data.frame(spec.fdr = numeric(),
                         tp.found.threshold = numeric(),
                         fp.found.threshold = numeric(),
                         tp.sim.threshold = numeric(),
                         num.regions = numeric(),
                         obs.fdr = numeric(),
                         power = numeric(),
                         total.length = numeric(),
                         method = character(),
                         stringsAsFactors = F)
  data.tmp.all <- data.frame(spec.fdr = numeric(),
                         tp.found.all = numeric(),
                         fp.found.all = numeric(),
                         tp.sim.all = numeric(),
                         num.regions = numeric(),
                         obs.fdr = numeric(),
                         power = numeric(),
                         total.length = numeric(),
                         method = character(),
                         stringsAsFactors = F)
  data.tmp.prop <- data.frame(spec.fdr = numeric(),
                         tp.found.prop = numeric(),
                         fp.found.prop = numeric(),
                         tp.sim.prop = numeric(),
                         num.regions = numeric(),
                         obs.fdr = numeric(),
                         power = numeric(),
                         total.length = numeric(),
                         method = character(),
                         stringsAsFactors = F)
  data.tmp.nuc <- data.frame(spec.fdr = numeric(),
                         tp.found.nuc = numeric(),
                         fp.found.nuc = numeric(),
                         tp.sim.nuc = numeric(),
                         num.regions = numeric(),
                         obs.fdr = numeric(),
                         power = numeric(),
                         total.length = numeric(),
                         method = character(),
                         stringsAsFactors = F)
  
  # for every FDR cut-off
  for (i in 1:length(cutoffs)) {
    fdr <- cutoffs[[i]]
    # number of detected regions
    num.regions <- length(ranges(sim.new[[i]][[j]]))
    # total length of detected regions
    tot.length <- sum(width(sim.new[[i]][[j]]))
    
    if (num.regions > 0) {
      # lengths of overlaps with simulated (true) DMRs
      overlaps.foundToSim <- unlist(lapply(c(1:num.regions), function(x) meta_GR_overlap(ranges(sim.new[[i]][[j]][x]), ranges(read.dmr))))
      proportions.foundToSim <- overlaps.foundToSim/width(sim.new[[i]][[j]])
      
      # any overlap
      tp.found.all <- sum(overlaps.foundToSim>0)
      fp.found.all <- num.regions - tp.found.all
      
      # overlap at least threshold
      tp.found.threshold <- sum(proportions.foundToSim >=threshold)
      fp.found.threshold <- num.regions - tp.found.threshold
      
      # proportion
      tp.found.prop <- sum(proportions.foundToSim)
      fp.found.prop <- num.regions - tp.found.prop
      
      # number of correctly classified nucleotides
      tp.found.nuc <- sum(overlaps.foundToSim)
      fp.found.nuc <- tot.length - tp.found.nuc
      
      # lengths of overlaps of simulated regions with the detected ones
      overlaps.simToFound <- unlist(lapply(c(1:length(read.dmr)), function(x) meta_GR_overlap(ranges(read.dmr)[x],  ranges(sim.new[[i]][[j]]))))
      proportions.simToFound <- overlaps.simToFound/width(read.dmr)
      
      # any overlap
      tp.sim.all <- sum(overlaps.simToFound > 0)
      
      # overlap at least threshold
      tp.sim.threshold <- sum(proportions.simToFound >= threshold)
      
      # proportion
      tp.sim.prop <- sum(proportions.simToFound)
      
      # nucleotide-wise
      tp.sim.nuc <- sum(overlaps.simToFound)
      
      # save the data
      data.tmp.threshold[i,] <- c(fdr, tp.found.threshold, fp.found.threshold, tp.sim.threshold, num.regions, fp.found.threshold/num.regions, tp.sim.threshold/100, tot.length, n)
      data.tmp.all[i,] <- c(fdr, tp.found.all, fp.found.all, tp.sim.all, num.regions, fp.found.all/num.regions, tp.sim.all/100, tot.length, n)
      data.tmp.prop[i,] <- c(fdr, tp.found.prop, fp.found.prop, tp.sim.prop, num.regions, fp.found.prop/num.regions, tp.sim.prop/100, tot.length, n)
      data.tmp.nuc[i,] <- c(fdr, tp.found.nuc, fp.found.nuc, tp.sim.nuc, tot.length, fp.found.nuc/tot.length, tp.sim.nuc/tot.length.sim, tot.length, n)
      
    } else {
      # no regions identified
      data.tmp.threshold[i,] <- c(fdr, 0, 0, 0, 0, 0, 0, 0, n)
      data.tmp.all[i,] <- c(fdr, 0, 0, 0, 0, 0, 0, 0, n)
      data.tmp.prop[i,] <- c(fdr, 0, 0, 0, 0, 0, 0, 0, n)
      data.tmp.nuc[i,] <- c(fdr, 0, 0, 0, 0, 0, 0, 0, n)
    }
  }
  
  stats.each.threshold[[j]] <- data.tmp.threshold
  names(stats.each.threshold)[j] <- n
  stats.each.all[[j]] <- data.tmp.all
  names(stats.each.all)[j] <- n
  stats.each.prop[[j]] <- data.tmp.prop
  names(stats.each.prop)[j] <- n
  stats.each.nuc[[j]] <- data.tmp.nuc
  names(stats.each.nuc)[j] <- n
}
```

For the methods which do not provide q-value estimation, only the observed FDR and power are computed:
```{r}
for (i in 1:length(read.sim.no.qval)) {
  n <- names(read.sim.no.qval)[i]
  # number of identified regions
  num.regions <- length(ranges(read.sim.no.qval[[i]]))
  # total length of identified regions
  tot.length <- sum(width(read.sim.no.qval[[i]]))
  
  data.tmp.threshold <- data.frame(spec.fdr = numeric(),
                         tp.found.threshold = numeric(),
                         fp.found.threshold = numeric(),
                         tp.sim.threshold = numeric(),
                         num.regions = numeric(),
                         obs.fdr = numeric(),
                         power = numeric(),
                         total.length = numeric(),
                         method = character(),
                         stringsAsFactors = F)
  data.tmp.all <- data.frame(spec.fdr = numeric(),
                         tp.found.all = numeric(),
                         fp.found.all = numeric(),
                         tp.sim.all = numeric(),
                         num.regions = numeric(),
                         obs.fdr = numeric(),
                         power = numeric(),
                         total.length = numeric(),
                         method = character(),
                         stringsAsFactors = F)
  data.tmp.prop <- data.frame(spec.fdr = numeric(),
                         tp.found.prop = numeric(),
                         fp.found.prop = numeric(),
                         tp.sim.prop = numeric(),
                         num.regions = numeric(),
                         obs.fdr = numeric(),
                         power = numeric(),
                         total.length = numeric(),
                         method = character(),
                         stringsAsFactors = F)
  data.tmp.nuc <- data.frame(spec.fdr = numeric(),
                         tp.found.nuc = numeric(),
                         fp.found.nuc = numeric(),
                         tp.sim.nuc = numeric(),
                         num.regions = numeric(),
                         obs.fdr = numeric(),
                         power = numeric(),
                         total.length = numeric(),
                         method = character(),
                         stringsAsFactors = F)
  
  # lengths of overlaps with the simulated (= true) regions
  overlaps.foundToSim <- unlist(lapply(c(1:num.regions), function(x) meta_GR_overlap(ranges(read.sim.no.qval[[i]][x]), ranges(read.dmr))))
  proportions.foundToSim <- overlaps.foundToSim/width(read.sim.no.qval[[i]])
  
  # any overlap    
  tp.found.all <- sum(overlaps.foundToSim>0)
  fp.found.all <- num.regions - tp.found.all
  
  # at least threshold    
  tp.found.threshold <- sum(proportions.foundToSim >=threshold)
  fp.found.threshold <- num.regions - tp.found.threshold
    
  # proportional  
  tp.found.prop <- sum(proportions.foundToSim)
  fp.found.prop <- num.regions - tp.found.prop
  
  # nucleotide-wise
  tp.found.nuc <- sum(overlaps.foundToSim)
  fp.found.nuc <- tot.length - tp.found.nuc
  

  # lengths of overlaps of the simulated regions with detected ones
  overlaps.simToFound <- unlist(lapply(c(1:length(read.dmr)), function(x) meta_GR_overlap(ranges(read.dmr)[x],  ranges(read.sim.no.qval[[i]]))))
  proportions.simToFound <- overlaps.simToFound/width(read.dmr)
     
  # any overlap 
  tp.sim.all <- sum(overlaps.simToFound > 0)
     
  # at least threshold 
  tp.sim.threshold <- sum(proportions.simToFound >= threshold)
    
  # proportional  
  tp.sim.prop <- sum(proportions.simToFound)
  
  # nucleotide-wise
  tp.sim.nuc <- sum(overlaps.simToFound)
  
  
  # save the data
  data.tmp.threshold[1,] <- c(NaN, tp.found.threshold, fp.found.threshold, tp.sim.threshold, num.regions, fp.found.threshold/num.regions, tp.sim.threshold/100, tot.length, n)
  data.tmp.all[1,] <- c(NaN, tp.found.all, fp.found.all, tp.sim.all, num.regions, fp.found.all/num.regions, tp.sim.all/100, tot.length, n)
  data.tmp.prop[1,] <- c(NaN, tp.found.prop, fp.found.prop, tp.sim.prop, num.regions, fp.found.prop/num.regions, tp.sim.prop/100, tot.length, n)
  data.tmp.nuc[1,] <- c(fdr, tp.found.nuc, fp.found.nuc, tp.sim.nuc, tot.length, fp.found.nuc/tot.length, tp.sim.nuc/tot.length.sim, tot.length, n)
  
  stats.each.threshold[[length(sim.new[[1]])+i]] <- data.tmp.threshold
  stats.each.all[[length(sim.new[[1]])+i]] <- data.tmp.all
  stats.each.prop[[length(sim.new[[1]])+i]] <- data.tmp.prop
  stats.each.nuc[[length(sim.new[[1]])+i]] <- data.tmp.nuc
}
```

Concatenate the data frames:
```{r}
data.all <- data.frame(rbindlist(stats.each.all))
for (i in 1:8) {
  data.all[,i] <- as.numeric(as.character(data.all[,i]))
}

data.prop <- data.frame(rbindlist(stats.each.prop))
for (i in 1:8) {
  data.prop[,i] <- as.numeric(as.character(data.prop[,i]))
}

data.threshold <- data.frame(rbindlist(stats.each.threshold))
for (i in 1:8) {
  data.threshold[,i] <- as.numeric(as.character(data.threshold[,i]))
}

data.nuc <- data.frame(rbindlist(stats.each.nuc))
for (i in 1:8) {
  data.nuc[,i] <- as.numeric(as.character(data.nuc[,i]))
}
```

# Plots

## Negative control

## Simulated data

### Observed FDR vs. power
The observed FDR vs. observed power for different cut-offs.
```{r}
ggplotly(ggplot(data.all, aes(x=obs.fdr, y=power, color=method, group=method)) + 
  geom_point() + 
  geom_line() +
  labs(x="observed FDR", y="observed power", title="any overlap") +
  theme_light() +
  xlim(0,1) + ylim(0,1))
  
ggplotly(ggplot(data.prop, aes(x=obs.fdr, y=power, color=method, group=method)) + 
  geom_point() + 
  geom_line() +
  labs(x="observed FDR", y="observed power", title="proportional") +
  theme_light() +
  xlim(0,1) + ylim(0,1))

ggplotly(ggplot(data.threshold, aes(x=obs.fdr, y=power, color=method, group=method)) + 
  geom_point() + 
  geom_line() +
  labs(x="observed FDR", y="observed power", title="overlap at least 80 %") +
  theme_light() +
  xlim(0,1) + ylim(0,1))
```
Since the relative ordering of the methods stays the same, the used method of classifying regions as TPs, FPs, etc. does not seem to play a big role. Thus, from now on we use only the data with any overlap.

### Specified FDR vs. observed FDR
How well do the methods control FDR?
```{r}
#only for any overlap
ggplotly(ggplot(data.all, aes(x=spec.fdr, y=obs.fdr, color=method, group=method)) + 
  geom_point() + 
  geom_line() +
  labs(x="specified FDR", y="observed FDR") +
  theme_light() +
  xlim(0,1) + ylim(0,1) +
  geom_abline(linetype=3))
```

### FP vs. TP
Number of false positives -- detected regions that have no overlap with a real DMR -- vs. number of true positives -- real DMRs which were discovered.
```{r}
ggplotly(ggplot(data.all, aes(x=fp.found.all, y=tp.sim.all, color=method, group=method)) + 
  geom_point() + 
  geom_line() +
  labs(x="found FPs", y="found TPs") +
  theme_light() )
```

### Average length of reported regions
Average length of reported region for FDR cut-off 0.05 (and average length of the regions reported by BSseq and MethPipe), compared to average length of the true DMRs.
```{r}
means <- data.frame(avg_length = numeric(),
  method = character(),
  stringsAsFactors = F)
# methods with qvals
for( i in 1:length(sim.new[[4]])) {
  mean <- mean(width(sim.new[[4]][[i]]))
  means[i,] <- c(mean, names(sim.new[[4]])[i])
}
# methods without qvals
for (i in 1:length(read.sim.no.qval)) {
  mean <- mean(width(read.sim.no.qval[[i]]))
  means[length(sim.new[[4]])+i,] <- c(mean, names(read.sim.no.qval)[i])
}
means$avg_length <- as.numeric(means$avg_length)

# mean length of the simulated DMRs
mean <- mean(width(read.dmr))

# plot
ggplotly(ggplot(means, aes(x=method, y=avg_length, fill=method)) +
           geom_bar(stat="identity") +
           theme_light() +
          labs(y="average length of reported region") +
          theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "none") +
          geom_hline(yintercept=mean, linetype="dashed") +
          geom_text(aes(3,mean+50,label = "average length of true DMRs", vjust = -1, size=3)))

```
The average length of ranges reported by `dmrseq` is the highest from all the methods (although it corresponds quite well with the average length of true DMRs). Isn't `dmrseq` good only because it reports longer intervals (that by chance overlap some true interval)? Let's check by looking at the individual nucleotides.

### FDR vs. power - nucleotide-wise
```{r}
ggplotly(ggplot(data.nuc, aes(x=obs.fdr, y=power, color=method, group=method)) + 
  geom_point() + 
  geom_line() +
  labs(x="observed FDR", y="observed power", title="nucleotide-wise") +
  theme_light() +
  xlim(0,1) + ylim(0,1))
```


# SessionInfo

```{r}
devtools::session_info()
```