---
bibliography: library.bib
csl: methods-in-ecology-and-evolution.csl
fontsize: 12pt
geometry: left=1in,right=1in,top=1.3in,bottom=1.3in
header-includes:
- \usepackage{placeins}
- \usepackage{graphics}
- \usepackage{fancyhdr}
- \usepackage{setspace}
- \usepackage{chngcntr}
- \usepackage{microtype}
- \usepackage{float}
- \usepackage{rotating}
- \onehalfspacing
- \counterwithin{figure}{section}
- \counterwithin{table}{section}
- \usepackage[T1]{fontenc}
- \usepackage{lmodern}
- \usepackage{mathptmx}
- \usepackage{cite}
- \usepackage[font=small,labelfont=bf]{caption}
- \usepackage[english]{babel}
- \usepackage{graphicx}
- \usepackage[fontsize=12pt]{scrextend}
linkcolor: black
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
  word_document:
    fig_caption: true
    toc: yes
    fig_width: 5
    fig_height: 5
urlcolor: black
citecolor: black
---

---
nocite: | 
  @2018
...

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.path = 'figures/',
                      echo = FALSE, warning = FALSE, message = FALSE)
```

\pagenumbering{gobble}

```{r child = '01_titlepage.Rmd'}
```


\newpage

\pagestyle{fancy}

\fancyhead[LE,RO]{}
\fancyhead[LO,RE]{}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0pt}




\pagenumbering{roman}

\fancyhead[CO,CE]{Abstract}
\section*{Abstract}
\addcontentsline{toc}{section}{Abstract}
\onehalfspacing

```{r child = '02_abstract.Rmd'}
```


\newpage

\fancyhead[CO,CE]{Acknowledgements}
\section*{Acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}
\onehalfspacing

```{r child = '03_acknowledgements.Rmd'}
```

\newpage
\fancyhead[CO,CE]{Table of Contents}
\setcounter{tocdepth}{10}
\tableofcontents

\newpage
\fancyhead[CO,CE]{List of Figures}
\addcontentsline{toc}{section}{List of Figures}
\listoffigures

\newpage
\fancyhead[CO,CE]{List of Tables}
\addcontentsline{toc}{section}{List of Tables}
\listoftables



\newpage

\fancyhead[CO,CE]{Abbreviations}
\section*{Abbreviations}
\addcontentsline{toc}{section}{Abbreviations}
\onehalfspacing

```{r child = '04_abbreviations.Rmd'}
```

\newpage

\pagenumbering{arabic}

\newpage
\fancyhead[CO,CE]{Introduction}
\onehalfspacing

```{r child = '05_introduction.Rmd'}
```


\FloatBarrier
\newpage
\fancyhead[CO,CE]{Methods}
\onehalfspacing

```{r child = '06_methods.Rmd'}
```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{Results}
\onehalfspacing

```{r child = '07_results.Rmd'}
```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{Conclusion}
\onehalfspacing

```{r child = '08_conclusion.Rmd'}
```

\newpage

\appendix
\setcounter{secnumdepth}{0}
\setcounter{figure}{0}
\setcounter{table}{0}
\renewcommand{\thefigure}{A.\arabic{figure}}
\renewcommand{\thetable}{A.\arabic{table}}

\fancyhead[CO,CE]{Appendix A: }
\section*{Appendix A: }
\addcontentsline{toc}{section}{Appendix A: }
\onehalfspacing

```{r child = '09_appendix1.Rmd'}
```


\newpage

\fancyhead[CO,CE]{References}
\addcontentsline{toc}{section}{References}
\onehalfspacing
```{r child = '10_references.Rmd'}
```