---
title: "bsseq"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

According to https://www.bioconductor.org/packages/devel/bioc/vignettes/bsseq/inst/doc/bsseq_analysis.html.

Setting directories:
```{r warning=FALSE, message=FALSE, eval=FALSE}
dataDir <- "data"
dir.create(file.path(dataDir))
resDir <- "results/bsseq"
dir.create(file.path(resDir))
```

#### Negative control
- read in data:
```{r warning=FALSE, message=FALSE, eval=FALSE}
library("bsseq")
library(SummarizedExperiment)
M <- read.table(paste(dataDir, "data_neg_control.M.gz", sep="/"), header=TRUE, sep="\t", stringsAsFactors = F)
cov <- read.table(paste(dataDir, "data_neg_control.cov.gz", sep="/"), header=TRUE, sep="\t", stringsAsFactors = F)
chr <- M$chr
pos <- M$pos
M$chr <- NULL
M$pos <- NULL
cov$chr <- NULL
cov$pos <- NULL
M <- data.matrix(M)
cov <- data.matrix(cov)
sampleNames <- gsub(".M", "", colnames(M))
colnames(M) <- sampleNames
colnames(cov) <- sampleNames

bs <- BSseq(chr=chr, pos=pos, M=M, Cov = cov, sampleNames=sampleNames)
grp1 <- c(1,3,5)
grp2 <- c(2,4,6)
```
- smoothing:
```{r warning=FALSE, message=FALSE, eval=FALSE}
bs.smoothed <- BSmooth(bs, verbose=TRUE, mc.cores = 16)
# load(paste(resDir,"bsseq_neg_control_smoothed.RData",sep="/"))
```

- we keep only loci with at least two samples in each group with coverage at least two:
```{r warning=FALSE, message=FALSE, eval=FALSE}
bs.cov <- getCoverage(bs.smoothed)
keepLoci.ex <- which(rowSums(bs.cov[, grp1] >= 5) >= 2 &
                     rowSums(bs.cov[, grp2] >= 5) >= 2)
length(keepLoci.ex)
bs.smoothed <- bs.smoothed[keepLoci.ex]
save(bs.smoothed, file=paste(resDir, "bsseq_neg_control_smoothed.RData",sep="/"))
```

- computing t-statistics:
```{r warning=FALSE, message=FALSE, eval=FALSE}
bs.tstat <- BSmooth.tstat(bs.smoothed, 
                                    group1 = grp1,
                                    group2 = grp2, 
                                    estimate.var = "same",
                                    local.correct = TRUE,
                                    verbose = TRUE)
plot(bs.tstat)
```

- DMRs identification and filtering (DMRs should have at least 10 CpGs and the mean difference should be at least 0.1; the first value is taken so as to be consistent with MethPipe; the mean difference is taken from the bsseq documentation):
```{r warning=FALSE, message=FALSE, eval=FALSE}
dmrs <- dmrFinder(bs.tstat)
dmrs.filtered <- subset(dmrs, n >= 10 & abs(meanDiff) >= 0.1)
nrow(dmrs.filtered)
write.table(dmrs.filtered, file=paste(resDir,"neg_control_bsseq_dmrs.filtered",sep="/"), quote=FALSE, sep="\t", col.names=FALSE)
```

- plotting:
```{r}
pData <- pData(bs.smoothed)
pData$col <- c("red", "blue", "red", "blue","red", "blue")
pData(bs.smoothed) <- pData
pdf(file = paste(resDir, "neg_control_bsseq_dmrs.pdf", sep="/"), width = 10, height = 5)
plotManyRegions(bs.smoothed, dmrs.filtered[1:6,], extend = 5000, 
                addRegions = dmrs.filtered)
dev.off()
```


#### Simulated data
```{r warning=FALSE, message=FALSE, eval=FALSE}
# read in data
M <- read.table(paste(dataDir, "sim_data.M.gz", sep="/"), header=TRUE, sep="\t", stringsAsFactors = F)
cov <- read.table(paste(dataDir, "sim_data.cov.gz", sep="/"), header=TRUE, sep="\t", stringsAsFactors = F)
chr <- M$chr.sim
pos <- M$pos.sim
M$chr.sim <- NULL
M$pos.sim <- NULL
cov$chr.sim <- NULL
cov$pos.sim <- NULL
M <- data.matrix(M)
cov <- data.matrix(cov)
sampleNames <- colnames(M)
bs <- BSseq(chr=chr, pos=pos, M=M, Cov = cov, sampleNames=sampleNames)
grp1 <- c(1,2,3)
grp2 <- c(4,5,6)

# smoothing
bs.smoothed <- BSmooth(bs, verbose=TRUE, mc.cores = 16)

# filtering regions with high enough coverage
bs.cov <- getCoverage(bs.smoothed)
keepLoci.ex <- which(rowSums(bs.cov[, grp1] >= 5) >= 2 &
                     rowSums(bs.cov[, grp2] >= 5) >= 2)
length(keepLoci.ex)
bs.smoothed <- bs.smoothed[keepLoci.ex]
save(bs.smoothed, file=paste(resDir, "bsseq_sim_data_smoothed.RData",sep="/"))

# computing t-test
bs.tstat <- BSmooth.tstat(bs.smoothed, 
                                    group1 = grp1,
                                    group2 = grp2, 
                                    estimate.var = "same",
                                    local.correct = TRUE,
                                    verbose = TRUE)
plot(bs.tstat)

# identifying and filtering DMRs
dmrs <- dmrFinder(bs.tstat)
dmrs.filtered <- subset(dmrs, n >= 10 & abs(meanDiff) >= 0.1)
nrow(dmrs.filtered)
write.table(dmrs.filtered, file=paste(resDir,"sim_data_bsseq_dmrs.filtered",sep="/"), quote=FALSE, sep="\t", col.names=FALSE)

# plotting
pData <- pData(bs.smoothed)
pData$col <- rep(c("red", "blue"), each=3)
pData(bs.smoothed) <- pData
pdf(file = paste(resDir, "sim_data_bsseq_dmrs.pdf", sep="/"), width = 10, height = 5)
plotManyRegions(bs.smoothed, dmrs.filtered[1:6,], extend = 5000, 
                addRegions = dmrs.filtered)
dev.off()
```


# Save Image
```{r}
save.image(file = "20181125_bsseq.RData", compress = T)
```

# SessionInfo

```{r}
devtools::session_info()
```