---
title: "Data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Negative control
- the same data as used in dmrseq paper (https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE64177)
- six samples of normal human dendritic cells (GSM1565940, GSM1565942, GSM1565944, GSM1565946, GSM1565948, GSM1565950)
- data restricted to chromosome 18, only positions present in all samples selected (using shell join), matrices M and cov created (shell cut)
- read in data and create `BSseq` object:
```{r message=FALSE, warning=FALSE, eval=FALSE}
library("dmrseq")
M <- read.csv("data/data_neg_control.M", header=TRUE, sep="\t")
cov <- read.csv("data/data_neg_control.cov", header=TRUE, sep="\t")
chr <- M$chr
pos <- M$pos
M$chr <- NULL
M$pos <- NULL
cov$chr <- NULL
cov$pos <- NULL
M <- data.matrix(M)
cov <- data.matrix(cov)
sampleNames <- gsub(".M", "", colnames(M))
colnames(M) <- sampleNames
colnames(cov) <- sampleNames

bs <- BSseq(chr=chr, pos=pos, M=M, Cov = cov, sampleNames=sampleNames)
```


### Simulated data
- 100 simulated DMRs added to negative control data
- simulated data created by `simDMRs` function from `dmrseq` package
```{r message=FALSE, warning=FALSE, eval=FALSE}
# reorder samples to create a null comparison 
bs.null <- bs[,c(1,3,5,2,4,6)]

# add 100 DMRs
bs.sim <- simDMRs(bs=bs.null, num.dmrs=100)

pos.sim <- start(bs.sim$bs)
chr.sim <- rep("chr18", n=length(pos.sim))
cov.sim <- cbind(chr.sim, pos.sim, getCoverage(bs.sim$bs, type="Cov"))
M.sim <- cbind(chr.sim, pos.sim, getCoverage(bs.sim$bs, type="M"))

write.table(cov.sim, file="data/sim_data.cov", sep="\t", quote=FALSE, row.names=FALSE)
write.table(M.sim, file="data/sim_data.M", sep="\t", quote=FALSE, row.names=FALSE)

ranges.start <- start(bs.sim$gr.dmrs)
ranges.end <- end(bs.sim$gr.dmrs)
mncov <- bs.sim$dmr.mncov
effect.sizes <- bs.sim$delta

sim.metadata <- data.frame(start=ranges.start, end=ranges.end, mean.cov=mncov, effect.size=effect.sizes)
write.table(sim.metadata, file="data/sim_data_meta.txt", sep="\t", quote=FALSE, row.names=FALSE)
```