---
title: "MethPipe"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Setting directories and path to MethPipe:
```{r warning=FALSE, message=FALSE, eval=FALSE}
dataDir <- "data"
dir.create(file.path(dataDir))
resDir <- "results/methpipe"
dir.create(file.path(resDir))
methPipe.path <- "../methpipe-3.4.3/bin"
bssseqDataDir <- "results/bsseq"
```

#### Negative control
Preparing the data in required format (see MethPipe documentation, page 6-7).
```{r warning=FALSE, message=FALSE, eval=FALSE}
M <- read.table(paste(dataDir, "data_neg_control.M.gz", sep="/"), header=TRUE, sep="\t", stringsAsFactors = F)
cov <- read.table(paste(dataDir, "data_neg_control.cov.gz", sep="/"), header=TRUE, sep="\t", stringsAsFactors = F)
chr <- M$chr
pos <- M$pos
M$chr <- NULL
M$pos <- NULL
cov$chr <- NULL
cov$pos <- NULL
grp1 <- c(1,3,5)
grp2 <- c(2,4,6)

# required coverage: at least 5 in at least 2 samples in each group
keepLoci.ex <- which(rowSums(cov[, grp1] >= 5) >= 2 &
                     rowSums(cov[, grp2] >= 5) >= 2)
length(keepLoci.ex)
cov <- cov[keepLoci.ex,]
M <- M[keepLoci.ex,]
chr <- chr[keepLoci.ex]
pos <- pos[keepLoci.ex]

strand <- rep('+', length(chr))  #we don't care about the strand, but methpipe requires it
context <- rep('CpG', length(chr)) #all data are from CpGs
for(i in 1:6){
  methCounts <- M[,i]
  covs <- cov[,i]
  methProp <- methCounts/covs
  data.i <- data.frame(chr=chr, pos=pos, strand=strand, context=context,
                       methProportion=methProp, coverage=covs)
  name <- paste(dataDir, "/data_neg_control_methpipe_", i, ".meth", sep="")
  write.table(data.i, file=name, sep="\t", quote=FALSE, col.names=FALSE, row.names=FALSE)
}
```

MethPipe analysis: First merging results from the replicates:
```{r warning=FALSE, message=FALSE, eval=FALSE}
cmd <- paste("./", methPipe.path, "/merge-methcounts ", dataDir, "/data_neg_control_methpipe_1.meth ", dataDir, "/data_neg_control_methpipe_3.meth ", dataDir, "/data_neg_control_methpipe_5.meth ", "-o ", dataDir, "/data_neg_control_methpipe_grp1.meth", sep="")
cmd
system(cmd)

cmd <- paste("./", methPipe.path, "/merge-methcounts ", dataDir, "/data_neg_control_methpipe_2.meth ", dataDir, "/data_neg_control_methpipe_4.meth ", dataDir, "/data_neg_control_methpipe_6.meth ", "-o ", dataDir, "/data_neg_control_methpipe_grp2.meth", sep="")
cmd
system(cmd)
```

Calculating the differential methylation score of each CpG site:
```{r warning=FALSE, message=FALSE, eval=FALSE}
cmd <- paste("./", methPipe.path, "/methdiff ", "-o ", resDir, "/neg_control_methpipe.methdiff ", dataDir, "/data_neg_control_methpipe_grp1.meth ", dataDir, "/data_neg_control_methpipe_grp2.meth", sep="")
cmd
system(cmd)
```

Identifying hypo- and hypermethylated regions (HMRs):
```{r warning=FALSE, message=FALSE, eval=FALSE}
cmd <- paste("./", methPipe.path, "/hmr ", "-o ", resDir, "/neg_control_grp1.hmr ", dataDir, "/data_neg_control_methpipe_grp1.meth", sep="")
cmd
system(cmd)

cmd <- paste("./", methPipe.path, "/hmr ", "-o ", resDir, "/neg_control_grp2.hmr ", dataDir, "/data_neg_control_methpipe_grp2.meth", sep="")
cmd
system(cmd)
```

Computing DMRs:
```{r warning=FALSE, message=FALSE, eval=FALSE}
cmd <- paste("./", methPipe.path, "/dmr ", resDir, "/neg_control_methpipe.methdiff ", resDir, "/neg_control_grp1.hmr ", resDir, "/neg_control_grp2.hmr ", resDir, "/neg_control_methpipe_dmrs_grp1 ", resDir, "/neg_control_methpipe_dmrs_grp2", sep="")
cmd
system(cmd)
```

Filtering only DMRs with at least 10 CpGs, from that at least 5 loci significant (as suggested by the documentation):
```{r warning=FALSE, message=FALSE, eval=FALSE}
cmd <- paste("awk -F '[:\t]' '$5 >= 10 && $6>=5 {print $0}' ", resDir, "/neg_control_methpipe_dmrs_grp1 > ", resDir, "/neg_control_methpipe_dmrs_grp1.filtered", sep="")
cmd
system(cmd)

cmd <- paste("awk -F '[:\t]' '$5 >= 10 && $6>=5 {print $0}' ", resDir, "/neg_control_methpipe_dmrs_grp2 > ", resDir, "/neg_control_methpipe_dmrs_grp2.filtered", sep="")
cmd
system(cmd)
```

Plotting using `bsseq` function `plotManyRegions`:
```{r warning=FALSE, message=FALSE, eval=FALSE}
# we need to load the bsseq smoothed data
load(paste(bssseqDataDir,"bsseq_neg_control_smoothed.RData",sep="/"))
pData <- pData(bs.smoothed)
pData$col <- rep(c("red", "blue"), nrep=3)
pData(bs.smoothed) <- pData

# load DMRs data
dmrs1 <- read.table(paste(resDir,"neg_control_methpipe_dmrs_grp1.filtered",sep="/"), header=FALSE, stringsAsFactors = FALSE, sep="\t")
dmrs1 <- dmrs1[,c(1,2,3)]
colnames(dmrs1) <- c("chr", "start", "end")

dmrs2 <- read.table(paste(resDir,"neg_control_methpipe_dmrs_grp2.filtered",sep="/"), header=FALSE, stringsAsFactors = FALSE, sep="\t")
dmrs2 <- dmrs2[,c(1,2,3)]
colnames(dmrs2) <- c("chr", "start", "end")

dmrs <- rbind(dmrs1, dmrs2)

# plotting
pdf(file = paste(resDir, "neg_control_methpipe_dmrs.pdf", sep="/"), width = 10, height = 5)
plotManyRegions(bs.smoothed, dmrs, extend = 5000, addRegions = dmrs)
dev.off()
```

####Simulated data
Preparation of data:
```{r warning=FALSE, message=FALSE, eval=FALSE}
M <- read.table(paste(dataDir, "/sim_data.M.gz", sep=""), header=TRUE, sep="\t", stringsAsFactors=F)
cov <- read.table(paste(dataDir, "/sim_data.cov.gz", sep=""), header=TRUE, sep="\t", stringsAsFactors=F)
chr <- M$chr.sim
pos <- M$pos.sim
M$chr.sim <- NULL
M$pos.sim <- NULL
cov$chr.sim <- NULL
cov$pos.sim <- NULL
# the groups are the same as in negative control, only the columns were shuffled during simulation
grp1 <- c(1,2,3)
grp2 <- c(4,5,6)

# required coverage: at least 5 in at least 2 samples in each group
keepLoci.ex <- which(rowSums(cov[, grp1] >= 5) >= 2 &
                     rowSums(cov[, grp2] >= 5) >= 2)
length(keepLoci.ex)
cov <- cov[keepLoci.ex,]
M <- M[keepLoci.ex,]
chr <- chr[keepLoci.ex]
pos <- pos[keepLoci.ex]

strand <- rep('+', length(chr))  #we don't care about the strand, but methpipe requires it
context <- rep('CpG', length(chr)) #all data are from CpGs
for(i in 1:6){
  methCounts <- M[,i]
  covs <- cov[,i]
  methProp <- methCounts/covs
  data.i <- data.frame(chr=chr, pos=pos, strand=strand, context=context,
                       methProportion=methProp, coverage=covs)
  name <- paste(dataDir, "/sim_data_methpipe_", i, ".meth", sep="")
  write.table(data.i, file=name, sep="\t", quote=FALSE, col.names=FALSE, row.names=FALSE)
}
```

MethPipe analysis:
```{r}
# merging data
cmd <- paste("./", methPipe.path, "/merge-methcounts ", dataDir, "/sim_data_methpipe_1.meth ", dataDir, "/sim_data_methpipe_2.meth ", dataDir, "/sim_data_methpipe_3.meth ", "-o ", dataDir, "/sim_data_methpipe_grp1.meth", sep="")
cmd
system(cmd)

cmd <- paste("./", methPipe.path, "/merge-methcounts ", dataDir, "/sim_data_methpipe_4.meth ", dataDir, "/sim_data_methpipe_5.meth ", dataDir, "/sim_data_methpipe_6.meth ", "-o ", dataDir, "/sim_data_methpipe_grp2.meth", sep="")
cmd
system(cmd)

# computing differential methylation score for each CpG
cmd <- paste("./", methPipe.path, "/methdiff ", "-o ", resDir, "/sim_data_methpipe.methdiff ", dataDir, "/sim_data_methpipe_grp1.meth ", dataDir, "/sim_data_methpipe_grp2.meth", sep="")
cmd
system(cmd)

# identifying HMRs
cmd <- paste("./", methPipe.path, "/hmr ", "-o ", resDir, "/sim_data_grp1.hmr ", dataDir, "/sim_data_methpipe_grp1.meth", sep="")
cmd
system(cmd)

cmd <- paste("./", methPipe.path, "/hmr ", "-o ", resDir, "/sim_data_grp2.hmr ", dataDir, "/sim_data_methpipe_grp2.meth", sep="")
cmd
system(cmd)

# identifying DMRs
cmd <- paste("./", methPipe.path, "/dmr ", resDir, "/sim_data_methpipe.methdiff ", resDir, "/sim_data_grp1.hmr ", resDir, "/sim_data_grp2.hmr ", resDir, "/sim_data_methpipe_dmrs_grp1 ", resDir, "/sim_data_methpipe_dmrs_grp2", sep="")
cmd
system(cmd)

# filtering DMRs
cmd <- paste("awk -F '[:\t]' '$5 >= 10 && $6 >= 5 {print $0}' ", resDir, "/sim_data_methpipe_dmrs_grp1 > ", resDir, "/sim_data_methpipe_dmrs_grp1.filtered", sep="")
cmd
system(cmd)

cmd <- paste("awk -F '[:\t]' '$5 >= 10 && $6 >= 5 {print $0}' ", resDir, "/sim_data_methpipe_dmrs_grp2 > ", resDir, "/sim_data_methpipe_dmrs_grp2.filtered", sep="")
cmd
system(cmd)

# plotting
load(paste(bssseqDataDir,"bsseq_sim_data_smoothed.RData",sep="/"))
pData <- pData(bs.smoothed)
pData$col <- rep(c("red", "blue"), each=3)
pData(bs.smoothed) <- pData

dmrs1 <- read.table(paste(resDir,"sim_data_methpipe_dmrs_grp1.filtered",sep="/"), header=FALSE, stringsAsFactors = FALSE, sep="\t")
dmrs1 <- dmrs1[,c(1,2,3)]
colnames(dmrs1) <- c("chr", "start", "end")

dmrs2 <- read.table(paste(resDir,"sim_data_methpipe_dmrs_grp2.filtered",sep="/"), header=FALSE, stringsAsFactors = FALSE, sep="\t")
dmrs2 <- dmrs2[,c(1,2,3)]
colnames(dmrs2) <- c("chr", "start", "end")

dmrs <- rbind(dmrs1, dmrs2)

pdf(file = paste(resDir, "sim_data_methpipe_dmrs.pdf", sep="/"), width = 10, height = 5)
plotManyRegions(bs.smoothed, dmrs, extend = 5000, addRegions = dmrs)
dev.off()
```