---
title: "MethylKit analysis"
author: "Deepak Tanwar"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    highlight: pyg
    keep_md: no
    number_sections: yes
    fig_width: 8.5
    fig_height: 11
    fig_caption: true
    df_print: paged
    code_folding: show
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  fontsize: 11pt
  geometry: margin=1in
  documentclass: article
  pdf_document:
    fig_caption: yes
    fig_height: 8
    fig_width: 8
    number_sections: yes
    toc: yes
    df_print: kable
    highlight: tango
  word_document: default
editor_options: 
  chunk_output_type: console
---

# Packages required

```{r}
library(methylKit)
library(data.table)
library(DT)
```

```{r, eval=FALSE, echo=FALSE}
# required coverage: at least 5 in at least 2 samples in each group
keepLoci.ex <- which(rowSums(cov[, grp1] >= 5) >= 2 &
                     rowSums(cov[, grp2] >= 5) >= 2)
```


# Negative control

## Input files
```{r}
files.neg <- list.files("./input", pattern = "neg", full.names = T)

files.list.neg <- as.list(files.neg)
names(files.list.neg) <- as.character(sapply(basename(files.neg), function(x) strsplit(x, "\\.")[[1]][1]))

myobj.neg <- methRead(
  location = files.list.neg, sample.id = as.list(names(files.list.neg)),
  assembly = "hg19", pipeline = "bismarkCoverage", header = F, skip = 0,
  dbtype = "tabix", treatment = c(0,1,0,1,0,1), dbdir = "./output/methylDB"
)
```


## Histogram of CpG methylation
```{r, fig.align='center', fig.height=11, fig.width=8.5}
par(mfrow = c(3, 2), cex = 0.6, cex.axis = 1.3, cex.lab = 1.3, cex.main = 1.2)
tmp <- sapply(1:length(files.list.neg), function(x)
  getMethylationStats(myobj.neg[[x]], plot = TRUE, both.strands = FALSE))
```


## Histogram of CpG coverage
```{r, fig.align='center', fig.height=11, fig.width=8.5}
par(mfrow = c(3, 2), cex = 0.6, cex.axis = 1.3, cex.lab = 1.3, cex.main = 1.2)
tmp <- sapply(1:length(files.list.neg), function(x)
  getCoverageStats(myobj.neg[[x]], plot = TRUE, both.strands = FALSE))
```


## Merging samples
```{r}
meth.neg <- unite(myobj.neg)
```


## Sample Correlation
```{r}
getCorrelation(meth.neg, plot = TRUE)
```


## Clustering samples
```{r}
clusterSamples(meth.neg, dist = "correlation", method = "ward", plot = TRUE)
```

```{r}
hc <- clusterSamples(meth.neg, dist = "correlation", method = "ward", plot = FALSE)
```


## PCA
```{r}
PCASamples(meth.neg, screeplot = TRUE)
PCASamples(meth.neg)
```


## Differential methylation
```{r}
myDiff.neg <- calculateDiffMeth(meth.neg, mc.cores = detectCores() - 1, overdispersion = "MN")
myDiff.neg.GR <- as(myDiff.neg,"GRanges")
myDiff.neg.GR.df <- data.frame(myDiff.neg.GR)

write.table(myDiff.neg.GR.df, file = gzfile("methylKit_negative_control.txt.gz", compression = 3),
            sep = "\t", row.names = F, quote = F)

tiles=tileMethylCounts(myobj.neg,win.size=1000,step.size=1000)
meth.neg.r <- unite(tiles)
myDiff.neg.r <- calculateDiffMeth(meth.neg.r, mc.cores = detectCores() - 1, overdispersion = "MN")
myDiff.neg.GR.r <- as(myDiff.neg.r,"GRanges")
myDiff.neg.GR.df.r <- data.frame(myDiff.neg.GR.r)

write.table(myDiff.neg.GR.df.r, file = gzfile("methylKit_negative_control_regions.txt.gz", 
                                              compression = 3),
            sep = "\t", row.names = F, quote = F)
```


# Simulated

## Input files
```{r}
files.sim <- list.files("./input", pattern = "sim", full.names = T)

files.list.sim <- as.list(files.sim)
names(files.list.sim) <- as.character(sapply(basename(files.sim), function(x) strsplit(x, "\\.")[[1]][1]))

myobj.sim <- methRead(
  location = files.list.sim, sample.id = as.list(names(files.list.sim)),
  assembly = "hg19", pipeline = "bismarkCoverage", header = F, skip = 0,
  dbtype = "tabix", treatment = c(0,0,0,1,1,1), dbdir = "./output/methylDB"
)
```


## Histogram of CpG methylation
```{r, fig.align='center', fig.height=11, fig.width=8.5}
par(mfrow = c(3, 2), cex = 0.6, cex.axis = 1.3, cex.lab = 1.3, cex.main = 1.2)
tmp <- sapply(1:length(files.list.sim), function(x)
  getMethylationStats(myobj.sim[[x]], plot = TRUE, both.strands = FALSE))
```


## Histogram of CpG coverage
```{r, fig.align='center', fig.height=11, fig.width=8.5}
par(mfrow = c(3, 2), cex = 0.6, cex.axis = 1.3, cex.lab = 1.3, cex.main = 1.2)
tmp <- sapply(1:length(files.list.sim), function(x)
  getCoverageStats(myobj.sim[[x]], plot = TRUE, both.strands = FALSE))
```


## Merging samples
```{r}
meth.sim <- unite(myobj.sim)
```


## Sample Correlation
```{r}
getCorrelation(meth.sim, plot = TRUE)
```


## Clustering samples
```{r}
clusterSamples(meth.sim, dist = "correlation", method = "ward", plot = TRUE)
```

```{r}
hc <- clusterSamples(meth.sim, dist = "correlation", method = "ward", plot = FALSE)
```


## PCA
```{r}
PCASamples(meth.sim, screeplot = TRUE)
PCASamples(meth.sim)
```


## Differential methylation
```{r}
myDiff.sim <- calculateDiffMeth(meth.sim, mc.cores = detectCores() - 1, overdispersion = "MN")
myDiff.sim.GR <- as(myDiff.sim,"GRanges")
myDiff.sim.GR.df <- data.frame(myDiff.sim.GR)

write.table(myDiff.sim.GR.df, file = gzfile("methylKit_sim_data.txt.gz", compression = 3),
            sep = "\t", row.names = F, quote = F)

tiles=tileMethylCounts(myobj.sim,win.size=1000,step.size=1000)
meth.sim.r <- unite(tiles)
myDiff.sim.r <- calculateDiffMeth(meth.sim.r, mc.cores = detectCores() - 1, overdispersion = "MN")
myDiff.sim.GR.r <- as(myDiff.sim.r,"GRanges")
myDiff.sim.GR.df.r <- data.frame(myDiff.sim.GR.r)

write.table(myDiff.sim.GR.df.r, file = gzfile("methylKit_sim_data_regions.txt.gz", 
                                              compression = 3),
            sep = "\t", row.names = F, quote = F)
```