---
title: "CGmapTools"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Method description
CGmapTools uses ‘dynamic-fragment finding method’. First, background fragments are defined using several criteria (contain a minimum number of CpGs, at most maximal distance between two adjacent CpGs, etc.). An unpaired t-test is then carried out to compare the methylation levels of CpG sites shared between fragments in the two groups. DMRs are then selected from these background fragments by P-values. 

#Analysis

Setting directories:
```{r warning=FALSE, message=FALSE, eval=FALSE}
dataDir <- "data"
dir.create(file.path(dataDir))
resDir <- "results/cgmaptools"
dir.create(file.path(resDir))
bssseqDataDir <- "results/bsseq"
```
CGmapTools must be installed and the directory added to `PATH`.

#### Negative control
- preparing data in the required format:
```{r warning=FALSE, message=FALSE, eval=FALSE}
M <- read.table(paste(dataDir, "data_neg_control.M.gz", sep="/"), header=TRUE, sep="\t", stringsAsFactors = F)
cov <- read.table(paste(dataDir, "data_neg_control.cov.gz", sep="/"), header=TRUE, sep="\t", stringsAsFactors = F)
chr <- M$chr
pos <- M$pos
M$chr <- NULL
M$pos <- NULL
cov$chr <- NULL
cov$pos <- NULL
grp1 <- c(1,3,5)
grp2 <- c(2,4,6)

# required coverage: at least 5 in at least 2 samples in each group
keepLoci.ex <- which(rowSums(cov[, grp1] >= 5) >= 2 &
                     rowSums(cov[, grp2] >= 5) >= 2)
length(keepLoci.ex)
cov <- cov[keepLoci.ex,]
M <- M[keepLoci.ex,]
chr <- chr[keepLoci.ex]
pos <- pos[keepLoci.ex]

context <- rep('CpG', length(chr)) #all data are from CpGs
Cs <- rep('C', length(chr)) #column of C's is part of the format

meth.1 <- rep(0, length(chr))
cov.1 <- rep(0, length(chr))

meth.2 <- rep(0, length(chr))
cov.2 <- rep(0, length(chr))
for(i in grp1){
  meth.1 <- meth.1 + M[,i]
  cov.1 <- cov.1 + cov[,i]
}
prop.1 <- meth.1/cov.1
for(i in grp2){
  meth.2 <- meth.2 + M[,i]
  cov.2 <- cov.2 + cov[,i]
}
prop.2 <- meth.2/cov.2

new.data <- data.frame(chr=chr,
                       C=Cs,
                       pos=pos,
                       context1=context,
                       context2=context,
                       prop.1=prop.1,
                       meth.1=meth.1,
                       cov.1=cov.1,
                       prop.2=prop.2,
                       meth.2=meth.2,
                       cov.2=cov.2)
gz1 <- gzfile(paste(dataDir,"data_neg_control_cgmaptools.txt.gz", sep="/"), "w")
write.table(new.data, file=gz1, sep="\t", quote=FALSE, col.names=FALSE, row.names=FALSE)
close(gz1)
```
- needed for plotting:
```{r warning=FALSE, message=FALSE, eval=FALSE}
# we need to load the bsseq smoothed data
load(paste(bssseqDataDir,"bsseq_neg_control_smoothed.RData",sep="/"))
pData <- pData(bs.smoothed)
pData$col <- rep(c("red", "blue"), nrep=3)
pData(bs.smoothed) <- pData
```


- run CGmapTools:
cgmaptools dmr -i data/data_neg_control_cgmaptools.txt.gz -n 10 -o results/cgmaptools/neg_control_dmrs.txt
```{r }
cmd <- paste("cgmaptools dmr -i ", dataDir, "/data_neg_control_cgmaptools.txt.gz -n 10 -o ", resDir, "/neg_control_cgmaptools_dmrs.txt",sep="")
cmd
system(cmd)
```

- filtering by Benjamini-Hochberg, for different values of FDR:
```{r warning=FALSE, message=FALSE, eval=FALSE}
dmrs <- read.table(paste(resDir, "/neg_control_cgmaptools_dmrs.txt",sep=""), header=F, sep="\t", stringsAsFactors = F)
p.values <- dmrs[,5]
p.values.adj <- p.adjust(p.values, method="BH")
for(fdr in seq(0.01,0.1,by=0.01)){
  dmrs.filtered <- dmrs[p.values.adj<=fdr,]
  dmrs.filtered[,1] <- gsub("b", "", dmrs.filtered[,1])
  write.table(dmrs.filtered, file=paste(resDir, "/neg_control_cgmaptools_dmrs_", fdr, ".filtered", sep=""), sep="\t", quote=FALSE, col.names=FALSE, row.names=FALSE)
    
  #sorting with respect to p-value
  dmrs.filtered.sorted <- dmrs.filtered[order(dmrs.filtered[,5]),]
  dmrs.filtered.sorted <- dmrs.filtered.sorted[,c(1,2,3)]
  colnames(dmrs.filtered.sorted) <- c("chr", "start", "end")
  
  # plotting
  n = min(6,nrow(dmrs.filtered.sorted))
  pdf(file = paste(resDir, "/neg_control_cgmaptools_dmrs_", fdr, ".pdf", sep=""), width = 10, height = 5)
  plotManyRegions(bs.smoothed, dmrs.filtered.sorted[1:n,], extend = 5000, addRegions = dmrs.filtered.sorted)
  dev.off()
}
```


#### Simulated data
- preparing data in the required format:
```{r warning=FALSE, message=FALSE, eval=FALSE}
M <- read.table(paste(dataDir, "/sim_data.M.gz", sep=""), header=TRUE, sep="\t", stringsAsFactors = F)
cov <- read.table(paste(dataDir, "/sim_data.cov.gz", sep=""), header=TRUE, sep="\t", stringsAsFactors = F)
chr <- M$chr.sim
pos <- M$pos.sim
M$chr.sim <- NULL
M$pos.sim <- NULL
cov$chr.sim <- NULL
cov$pos.sim <- NULL
grp1 <- c(1,2,3)
grp2 <- c(4,5,6)

# required coverage: at least 5 in at least 2 samples in each group
keepLoci.ex <- which(rowSums(cov[, grp1] >= 5) >= 2 &
                     rowSums(cov[, grp2] >= 5) >= 2)
length(keepLoci.ex)
cov <- cov[keepLoci.ex,]
M <- M[keepLoci.ex,]
chr <- chr[keepLoci.ex]
pos <- pos[keepLoci.ex]

context <- rep('CpG', length(chr)) #all data are from CpGs
Cs <- rep('C', length(chr)) #column of C's is part of the format

meth.1 <- rep(0, length(chr))
cov.1 <- rep(0, length(chr))

meth.2 <- rep(0, length(chr))
cov.2 <- rep(0, length(chr))
for(i in grp1){
  meth.1 <- meth.1 + M[,i]
  cov.1 <- cov.1 + cov[,i]
}
prop.1 <- meth.1/cov.1
for(i in grp2){
  meth.2 <- meth.2 + M[,i]
  cov.2 <- cov.2 + cov[,i]
}
prop.2 <- meth.2/cov.2

new.data <- data.frame(chr=chr,
                       C=Cs,
                       pos=pos,
                       context1=context,
                       context2=context,
                       prop.1=prop.1,
                       meth.1=meth.1,
                       cov.1=cov.1,
                       prop.2=prop.2,
                       meth.2=meth.2,
                       cov.2=cov.2)
gz1 <- gzfile(paste(dataDir,"sim_data_cgmaptools.txt.gz", sep="/"), "w")
write.table(new.data, file=gz1, sep="\t", quote=FALSE, col.names=FALSE, row.names=FALSE)
close(gz1)
```

- needed for plotting:
```{r warning=FALSE, message=FALSE, eval=FALSE}
# we need to load the bsseq smoothed data
load(paste(bssseqDataDir,"bsseq_sim_data_smoothed.RData",sep="/"))
pData <- pData(bs.smoothed)
pData$col <- rep(c("red", "blue"), each=3)
pData(bs.smoothed) <- pData
```

- run CGmapTools:
```{r warning=FALSE, message=FALSE, eval=FALSE}
cmd <- paste("cgmaptools dmr -i ", dataDir, "/sim_data_cgmaptools.txt.gz -n 10 -o ", resDir, "/sim_data_cgmaptools_dmrs.txt",sep="")
cmd
system(cmd)
```

- filter and plot:
```{r warning=FALSE, message=FALSE, eval=FALSE}
dmrs <- read.table(paste(resDir, "/sim_data_cgmaptools_dmrs.txt",sep=""), header=F, sep="\t", stringsAsFactors = F)
p.values <- dmrs[,5]
p.values.adj <- p.adjust(p.values, method="BH")
for(fdr in seq(0.01,0.1,by=0.01)){
  dmrs.filtered <- dmrs[p.values.adj<=fdr,]
  dmrs.filtered[,1] <- gsub("b", "", dmrs.filtered[,1])
  print(nrow(dmrs.filtered))
  write.table(dmrs.filtered, file=paste(resDir, "/sim_data_cgmaptools_dmrs_", fdr, ".filtered", sep=""), sep="\t", quote=FALSE, col.names=FALSE, row.names=FALSE)
    
  #sorting with respect to p-value
  dmrs.filtered.sorted <- dmrs.filtered[order(dmrs.filtered[,5]),]
  dmrs.filtered.sorted <- dmrs.filtered.sorted[,c(1,2,3)]
  colnames(dmrs.filtered.sorted) <- c("chr", "start", "end")
  
  # plotting
  n = min(6,nrow(dmrs.filtered.sorted))
  pdf(file = paste(resDir, "/sim_data_cgmaptools_dmrs_", fdr, ".pdf", sep=""), width = 10, height = 5)
  plotManyRegions(bs.smoothed, dmrs.filtered.sorted[1:n,], extend = 5000, addRegions = dmrs.filtered.sorted)
  dev.off()
}
```
