---
title: "DMRs: CGmapTools"
author: "Hana Parizkova & Deepak Tanwar"
date: "<b>Created on:</b> 2018-11-29 <br> <b>Updated on:</b> `r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    highlight: pyg
    keep_md: no
    number_sections: yes
    fig_width: 8.5
    fig_height: 11
    fig_caption: true
    df_print: paged
    code_folding: show
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  fontsize: 12pt
  geometry: margin=1in
  documentclass: article
---

# Packages required
```{r, warning=F, message=FALSE}
library(DT)
library(data.table)
```


# Merge groups into one table
First, the data from each group need to be merged into a single file.
## Negative control

### Group1
```{r, engine='bash'}
mkdir -p ./output/01_merged_tables

cgmaptools mergelist tosingle -i <(zcat ./input/NC_rep1.cov.gz),<(zcat ./input/NC_rep3.cov.gz),<(zcat ./input/NC_rep5.cov.gz) | pigz -11 -p 16 > ./output/01_merged_tables/neg_control_grp1.merge.gz
```

### Group2
```{r, engine='bash'}
cgmaptools mergelist tosingle -i <(zcat ./input/NC_rep2.cov.gz),<(zcat ./input/NC_rep4.cov.gz),<(zcat ./input/NC_rep6.cov.gz) | pigz -11 -p 16 > ./output/01_merged_tables/neg_control_grp2.merge.gz
```

## Simulated data

### Group1
```{r, engine='bash'}
cgmaptools mergelist tosingle -i <(zcat ./input/sim_rep1.cov.gz),<(zcat ./input/sim_rep3.cov.gz),<(zcat ./input/sim_rep5.cov.gz) | pigz -11 -p 16 > ./output/01_merged_tables/sim_data_grp1.merge.gz
```

### Group2
```{r, engine='bash'}
cgmaptools mergelist tosingle -i <(zcat ./input/sim_rep2.cov.gz),<(zcat ./input/sim_rep4.cov.gz),<(zcat ./input/sim_rep6.cov.gz) | pigz -11 -p 16 > ./output/01_merged_tables/sim_data_grp2.merge.gz
```


# Tables to perform DMRs analysis
The files with data from group 1 and group 2 are then also merged using `intersect` function.

## Negative control
```{r, engine='bash'}
mkdir -p ./output/02_intersect_grp1_grp2

cgmaptools intersect -1 <(zcat ./output/01_merged_tables/neg_control_grp1.merge.gz) -2 <(zcat ./output/01_merged_tables/neg_control_grp2.merge.gz) | pigz -11 -p 16 > ./output/02_intersect_grp1_grp2/neg_control.intersect.gz
```

## Simulated data
```{r, engine='bash'}
cgmaptools intersect -1 <(zcat ./output/01_merged_tables/sim_data_grp1.merge.gz) -2 <(zcat ./output/01_merged_tables/sim_data_grp2.merge.gz) | pigz -11 -p 16 > ./output/02_intersect_grp1_grp2/sim_data.intersect.gz
```


# DMRs analysis
We compute DMRs.
## Negative Control
```{r, engine='bash'}
mkdir -p ./output/03_dmr

cgmaptools dmr -i <(zcat ./output/02_intersect_grp1_grp2/neg_control.intersect.gz) -n 10 | pigz -11 -p 16 > ./output/03_dmr/neg_control.dmr.gz
```

## Simulated data
```{r, engine='bash'}
cgmaptools dmr -i <(zcat ./output/02_intersect_grp1_grp2/sim_data.intersect.gz) -n 10 | pigz -11 -p 16 > ./output/03_dmr/sim_data.dmr.gz
```


# DMR Tables
The provided p-values are adjusted for multiple testing by Benjamini-Hochberg correction.
## Negative control

```{r}
dmr.nc <- read.table("./output/03_dmr/neg_control.dmr.gz",
  header = F, sep = "\t", stringsAsFactors = F,
  col.names = c("chr", "start", "end", "stat", "pVal", "mGrp1", "mGrp2", "N_site")
)

# adjust p-values
dmr.nc$qval <- p.adjust(p = dmr.nc$pVal, method = "BH")
dmr.nc <- dmr.nc[order(dmr.nc$qval), ]
# select only regions with methylation difference at least 0.1
dmr.nc <- dmr.nc[abs(dmr.nc$mGrp1-dmr.nc$mGrp2) >= 0.1,]

write.table(dmr.nc, file = gzfile("./output/cgmaptools_dmr_neg_control.txt.gz"), sep = "\t", quote = F, row.names = F)
```

```{r}
c1 <- dmr.nc$qval[dmr.nc$qval <= 0.05]
datatable(
  dmr.nc,
  rownames = F,
  filter = "top", extensions = c("Buttons", "ColReorder"), options = list(
    pageLength = 10,
    buttons = c("copy", "csv", "excel", "pdf", "print"),
    colReorder = list(realtime = FALSE),
    dom = "fltBip"
  )
) %>%
  formatStyle(
    "qval",
    backgroundColor = styleEqual(c1, rep("#66C2A5", length(c1)))
  )
```



## Simulated data
```{r}
dmr.sim <- read.table("./output/03_dmr/sim_data.dmr.gz",
  header = F, sep = "\t", stringsAsFactors = F,
  col.names = c("chr", "start", "end", "t", "pVal", "mGrp1", "mGrp2", "N_site")
)

# adjust p-values
dmr.sim$qval <- p.adjust(p = dmr.sim$pVal, method = "BH")
dmr.sim <- dmr.sim[order(dmr.sim$qval), ]
# select only regions with methylation difference at least 0.1
dmr.sim <- dmr.sim[abs(dmr.sim$mGrp1-dmr.sim$mGrp2) >= 0.1,]
write.table(dmr.sim, file = gzfile("./output/cgmaptools_dmr_sim_data.txt.gz"), sep = "\t", quote = F, row.names = F)
```


```{r}
c2 <- dmr.sim$qval[dmr.sim$qval <= 0.05]
datatable(
  dmr.sim,
  rownames = F,
  filter = "top", extensions = c("Buttons", "ColReorder"), options = list(
    pageLength = 10,
    buttons = c("copy", "csv", "excel", "pdf", "print"),
    colReorder = list(realtime = FALSE),
    dom = "fltBip"
  )
) %>%
  formatStyle(
    "qval",
    backgroundColor = styleEqual(c2, rep("#66C2A5", length(c2)))
  )
```


# SessionInfo

```{r}
devtools::session_info()
```