---
title: "Coverage plots"
author: "Deepak Tanwar & Hana Parizkova"
date: "<b>Created on:</b> 2019-01-08 <br> <b>Updated on:</b> `r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    highlight: pyg
    keep_md: no
    number_sections: yes
    fig_width: 10
    fig_height: 7
    fig_caption: true
    df_print: paged
    code_folding: show
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  fontsize: 11pt
  geometry: margin=1in
  documentclass: article
---

# Libraries

```{r, message=FALSE, warning=FALSE}
library(data.table)
library(edgeR)
library(reshape2)
library(plotly)
library(ggdendro)
library(viridis)
library(autoplotly)
library(ggfortify)
library(ClusterR)
```


# Read Bismark files

## Negative control
```{r}
files.nc <- list.files(path = "./input", pattern = "NC_rep", full.names = T)

anno.nc <- read.table("./input/anno_neg_control.txt.gz", header = T, sep = "\t", stringsAsFactors = F)

bismark.nc <- readBismark2DGE(files = files.nc, sample.names = anno.nc$names)
```


## Simulated data
```{r}
files.sim <- list.files(path = "./input", pattern = "sim_rep", full.names = T)

anno.sim <- read.table("./input/anno_sim_data.txt.gz", header = T, sep = "\t", stringsAsFactors = F)

bismark.sim <- readBismark2DGE(files = files.sim, sample.names = anno.sim$names)
```



# Filtering and plotting

## Function
```{r}
beta_calc <- function(objectDGE, coverageThreshold, anno = anno) {
  ## Function to calculate Beta values from DGE object obtained from readBismark2DGE function.
  ## objectDGE: DGEList
  ## coverageThreshold: Coverage threshold to be used for all samples
  ## Return: Coverage, Coverage melted table, distribution, Beta value table, Beta value melted table, distribution

  # Counts (raw coverage)
  df <- data.frame(objectDGE$counts, stringsAsFactors = F, check.names = F)

  me.raw <- df[, grep("Me", colnames(df))]
  un.raw <- df[, grep("Un", colnames(df))]
  cov.raw <- me.raw + un.raw


  # Filter based on coverage
  data <- NULL

  if (is.null(coverageThreshold)) {
    data <- df
    coverageThreshold <- "NO filtering"
  } else {
    data <- df[rowSums(cov.raw > coverageThreshold) == ncol(cov.raw), ]
  }

  # Beta valus after filtering
  me <- data[, grep("Me", colnames(data))]
  un <- data[, grep("Un", colnames(data))]
  beta <- me / (me + un)
  colnames(beta) <- as.character(sapply(colnames(beta), function(x) strsplit(x, "-")[[1]][1]))

  # Beta values for plotting
  reshape.beta <- data.frame(t(beta))
  reshape.beta$Samples <- rownames(reshape.beta)
  reshape.beta.melt <- melt(reshape.beta)

  # Plotting Beta value
  p1 <- ggplotly(ggplot(reshape.beta.melt, aes(x = value)) +
    geom_histogram(aes(y = ..density..), alpha = 0.7, fill = "#333333") +
    geom_density(fill = "#CC6666", alpha = 0.3) +
    theme(panel.background = element_rect(fill = "#ffffff")) +
    facet_wrap(~Samples, nrow = 3, shrink = T) +
    ggtitle(paste0("Beta distribution (coverage > ", coverageThreshold, ")")) +
    labs(x = "Beta value", y = ""))

  # Coverage plot after filtering
  cov <- me + un
  colnames(cov) <- as.character(sapply(colnames(cov), function(x) strsplit(x, "-")[[1]][1]))
  reshape.cov <- data.frame(t(cov))
  reshape.cov$Samples <- rownames(reshape.cov)
  reshape.cov.melt <- melt(reshape.cov)
  reshape.cov.melt.log2 <- reshape.cov.melt
  reshape.cov.melt.log2$value <- log2(reshape.cov.melt.log2$value)

  p2 <- ggplotly(ggplot(reshape.cov.melt.log2, aes(x = value)) +
    geom_histogram(aes(y = ..density..), alpha = 0.7, fill = "#333333") +
    geom_density(fill = "#20AAEA", alpha = 0.3) +
    theme(panel.background = element_rect(fill = "#ffffff")) +
    facet_wrap(~Samples, nrow = 3, shrink = T) +
    ggtitle(paste0("Coverage distribution (coverage > ", coverageThreshold, ")")) +
    labs(x = "Coverage (log2 scaled)", y = ""))

  # PCA plot

  ## Coverage
  pca.cov <- prcomp(t(cov))

  p3.cov <- plot_ly(
    x = pca.cov$x[, 1], y = pca.cov$x[, 2],
    data = anno, color = ~Group, mode = "markers",
    symbol = ~ as.numeric(as.factor(Group)),
    marker = list(size = 10), hoverinfo = "text",
    text = ~ paste("<b>Group:</b> ", Group, "<br><b>SampleID:</b> ", names),
    type = "scatter"
  ) %>%
    layout(
      title = paste0(
        "<b>Coverage PCA plot (<span style='color: blue;'>coverage > ",
        coverageThreshold, "</span>)<b>"
      ),
      xaxis = list(title = "<b><i>PC1<i><b>"),
      yaxis = list(title = "<b><i>PC2<i><b>")
    )


  ## Beta
  pca.beta <- prcomp(t(beta))

  p3.beta <- plot_ly(
    x = pca.beta$x[, 1], y = pca.beta$x[, 2],
    data = anno, color = ~Group, mode = "markers",
    symbol = ~ as.numeric(as.factor(Group)),
    marker = list(size = 10), hoverinfo = "text",
    text = ~ paste("<b>Group:</b> ", Group, "<br><b>SampleID:</b> ", names),
    type = "scatter"
  ) %>%
    layout(
      title = paste0(
        "<b>Beta PCA plot (<span style='color: blue;'>coverage > ",
        coverageThreshold, "</span>)<b>"
      ),
      xaxis = list(title = "<b><i>PC1<i><b>"),
      yaxis = list(title = "<b><i>PC2<i><b>")
    )


  # MDS plot

  ## Coverage
  mds.cov <- cmdscale(dist(t(cov)))

  p4.cov <- plot_ly(
    x = mds.cov[, 1], y = mds.cov[, 2],
    data = anno, color = ~Group, mode = "markers",
    symbol = ~ as.numeric(as.factor(Group)),
    marker = list(size = 10), hoverinfo = "text",
    text = ~ paste("<b>Group:</b> ", Group, "<br><b>SampleID:</b> ", names),
    type = "scatter"
  ) %>%
    layout(
      title = paste0(
        "<b>Coverage MDS plot (<span style='color: blue;'>coverage > ",
        coverageThreshold, "</span>)<b>"
      ),
      xaxis = list(title = "<b><i>MDS1<i><b>"),
      yaxis = list(title = "<b><i>MDS2<i><b>")
    )


  ## Beta
  mds.beta <- cmdscale(dist(t(beta)))

  p4.beta <- plot_ly(
    x = mds.beta[, 1], y = mds.beta[, 2],
    data = anno, color = ~Group, mode = "markers",
    symbol = ~ as.numeric(as.factor(Group)),
    marker = list(size = 10), hoverinfo = "text",
    text = ~ paste("<b>Group:</b> ", Group, "<br><b>SampleID:</b> ", names),
    type = "scatter"
  ) %>%
    layout(
      title = paste0(
        "<b>Beta MDS plot (<span style='color: blue;'>coverage > ",
        coverageThreshold, "</span>)<b>"
      ),
      xaxis = list(title = "<b><i>MDS1<i><b>"),
      yaxis = list(title = "<b><i>MDS2<i><b>")
    )

  # SVD

  ## Coverage

  svd.cov <- svd(x = t(cov))

  p5.cov <- plot_ly(
    x = svd.cov$u[, 1], y = svd.cov$u[, 2],
    data = anno, color = ~Group, mode = "markers",
    symbol = ~ as.numeric(as.factor(Group)),
    marker = list(size = 10), hoverinfo = "text",
    text = ~ paste("<b>Group:</b> ", Group, "<br><b>SampleID:</b> ", names),
    type = "scatter"
  ) %>%
    layout(
      title = paste0(
        "<b>Coverage SVD plot (<span style='color: blue;'>coverage > ",
        coverageThreshold, "</span>)<b>"
      ),
      xaxis = list(title = "<b><i>SVD1<i><b>"),
      yaxis = list(title = "<b><i>SVD2<i><b>")
    )

  ## Beta
  svd.beta <- svd(x = t(beta))


  p5.beta <- plot_ly(
    x = svd.beta$u[, 1], y = svd.beta$u[, 2],
    data = anno, color = ~Group, mode = "markers",
    symbol = ~ as.numeric(as.factor(Group)),
    marker = list(size = 10), hoverinfo = "text",
    text = ~ paste("<b>Group:</b> ", Group, "<br><b>SampleID:</b> ", names),
    type = "scatter"
  ) %>%
    layout(
      title = paste0(
        "<b>Beta SVD plot (<span style='color: blue;'>coverage > ",
        coverageThreshold, "</span>)<b>"
      ),
      xaxis = list(title = "<b><i>SVD1<i><b>"),
      yaxis = list(title = "<b><i>SVD2<i><b>")
    )


  # Clustering

  ## Coverage
  p6.cov <- ggdendrogram(hclust(dist(t(cov)))) +
    ggtitle(paste0("Coverage dendrogram (coverage > ", coverageThreshold, ")"))

  ## Beta
  p6.beta <- ggdendrogram(hclust(dist(t(beta)))) +
    ggtitle(paste0("Beta dendrogram (coverage > ", coverageThreshold, ")"))


  list <- list(
    cov, reshape.cov.melt, p2, beta, reshape.beta.melt, p1,
    p3.cov, p4.cov, p5.cov, p6.cov, p3.beta, p4.beta, p5.beta, p6.beta
  )
  names(list) <- c(
    "Coverage", "reshapedCoverage", "DHPlotCov", "BetaValues", "reshapedBetaValues", "DHplotBeta",
    "PCAcov", "MDScov", "SVDcov", "HCcov", "PCAbeta", "MDSbeta", "SVDbeta", "HCbeta"
  )

  return(list)
}
```


## Calculation

### Negative control
```{r, warning=F, message=F}
cov.nc <- beta_calc(objectDGE = bismark.nc, coverageThreshold = NULL, anno = anno.nc)
```


### Simulated data
```{r, warning=F, message=F}
cov.sim <- beta_calc(objectDGE = bismark.sim, coverageThreshold = NULL, anno = anno.sim)
```


# Coverage Plots {.tabset .tabset-pills}

## Negative control

```{r, warning=F, message=F}
cov.nc$DHPlotCov
```


## Simulated data

```{r, warning=F, message=F}
cov.sim$DHPlotCov
```


# Beta Plots {.tabset .tabset-pills}

## Negative control

```{r, warning=F, message=F}
cov.nc$DHplotBeta
```

## Simulated data

```{r, warning=F, message=F}
cov.sim$DHplotBeta
```


# Clustering of samples {.tabset .tabset-pills}

## Negative control
```{r, warning=F, message=F}
cov.nc$HCcov
cov.nc$HCbeta
```


## Simulated data
```{r, warning=F, message=F}
cov.sim$HCcov
cov.sim$HCbeta
```


# PCA Plots {.tabset .tabset-pills}

## Negative control

```{r, warning=F, message=F}
cov.nc$PCAcov
cov.nc$PCAbeta
```


## Simulated data

```{r, warning=F, message=F}
cov.sim$PCAcov
cov.sim$PCAbeta
```



# MDS Plots {.tabset .tabset-pills}

## Negative control

```{r, warning=F, message=F}
cov.nc$MDScov
cov.nc$MDSbeta
```


## Simulated data

```{r, warning=F, message=F}
cov.sim$MDScov
cov.sim$MDSbeta
```



# SVD Plots {.tabset .tabset-pills}

## Negative control

```{r, warning=F, message=F}
cov.nc$SVDcov
cov.nc$SVDbeta
```


## Simulated data

```{r, warning=F, message=F}
cov.sim$SVDcov
cov.sim$SVDbeta
```


# SessionInfo
```{r}
devtools::session_info()
```
