---
title: "pWGBSSimla: BSseq DMRs"
author: "Deepak Tanwar"
date: "<b>Created on:</b> 2020-02-14 <br> <b>Updated on:</b> `r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: tango
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    keep_md: no
    number_sections: no
    fig_width: 8
    fig_height: 8
    fig_caption: true
    df_print: paged
    code_folding: show
  fontsize: 12pt
  geometry: margin=1in
  documentclass: article
# bibliography: references.bib
link-citations: yes
---

Analysis following the [MethPipe manual](http://smithlabresearch.org/downloads/methpipe-manual.pdf).

# Packages required
```{r lib, warning=F, message=FALSE}
library(DT)
library(plyr)
```

# Method description
Differential methylation scores are first computed at individual CpG sites in a way similar to one-directional version of Fisher’s test. HMRs (hypomethylated regions) are detected using Hidden Markov Model individually in each sample (group). Non-overlapping intervals of HMRs are then checked if they contain sufficient number of DMLs.


# Merging results from the replicates
The methylation data within each group first need to be combined into a single file using `merge-methcounts`.

## Simulated data

### Group1
```{r, engine='bash'}
mkdir -p ./output/01_merge_methcounts

merge-methcounts <(zcat ./input/sim_rep1.cov.gz) <(zcat ./input/sim_rep2.cov.gz) <(zcat ./input/sim_rep3.cov.gz) <(zcat ./input/sim_rep4.cov.gz) <(zcat ./input/sim_rep5.cov.gz) | pigz -11 -p 16 > ./output/01_merge_methcounts/sim_data_grp1.meth.gz
```

### Group2
```{r, engine='bash'}
merge-methcounts <(zcat ./input/sim_rep6.cov.gz) <(zcat ./input/sim_rep7.cov.gz) <(zcat ./input/sim_rep8.cov.gz) <(zcat ./input/sim_rep9.cov.gz) <(zcat ./input/sim_rep10.cov.gz) | pigz -11 -p 16 > ./output/01_merge_methcounts/sim_data_grp2.meth.gz
```


# Differential methylation score of each CpG site
Differential methylation score of each CpG site is computed. This gives us probability that the methylation level at the particular CpG site is lower in the first group than in the second.

## Simulated data
```{r, engine='bash'}
mkdir -p ./output/02_methdiff

methdiff <(zcat ./output/01_merge_methcounts/sim_data_grp1.meth.gz) <(zcat ./output/01_merge_methcounts/sim_data_grp2.meth.gz) | pigz -11 -p 16 > ./output/02_methdiff/sim_data.methdiff.gz
```


# Identifying hypomethylated regions (HMRs)
Hypomethylated regions (HMRs) need to be computed.


## Simulated data

### Group1
```{r, engine='bash'}
mkdir -p ./output/03_hmr

hmr <(zcat ./output/01_merge_methcounts/sim_data_grp1.meth.gz) | pigz -11 -p 16 > ./output/03_hmr/sim_data_grp1.hmr.gz
```

### Group2
```{r, engine='bash'}
hmr <(zcat ./output/01_merge_methcounts/neg_control_grp2.meth.gz) | pigz -11 -p 16 > ./output/03_hmr/sim_data_grp2.hmr.gz
```


# Computing DMRs
Finally, DMRs can be detected using the `methdiff` and `hmr` files.

## Simulated data

```{r, engine='bash'}
mkdir -p ./output/04_dmr

dmr <(zcat ./output/02_methdiff/sim_data.methdiff.gz) <(zcat ./output/03_hmr/sim_data_grp1.hmr.gz) <(zcat ./output/03_hmr/sim_data_grp2.hmr.gz) ./output/04_dmr/sim_data_grp1.dmr ./output/04_dmr/sim_data_grp2.dmr

pigz -11 -p 16 ./output/04_dmr/*.dmr
```


# Filtering
MethPipe does not provide the user with the effect size in the identified region, it only provides the total number of CpGs within the region and how many of them were significantly differently methylated. Thus, as suggested by the documentation, we have decided to filter out only those regions that are at least 10 CpGs long and provide at least 5 significantly differently methylated CpGs.

```{r, engine='bash'}
mkdir -p ./output/05_fltered_dmr

for i in ./output/04_dmr/*.dmr.gz
  do
  bname=`basename $i .dmr.gz`
  fname=./output/05_fltered_dmr/${bname}.filt.dmr.gz
  
  awk -F '[:\t]' '$5 >= 10 && $6>=5 {print $0}' <(zcat $i) | pigz -11 -p 16 > $fname
done
```



# DMR Tables

## Simulated data
```{r}
files.sim <- list.files("./output/05_fltered_dmr", pattern = "sim", full.names = T)
dmr.sim <- lapply(files.sim, function(x) read.table(x,
    sep = "\t", header = F, stringsAsFactors = F,
    col.names = c("chr", "start", "end", "nCpG", "sig.CpG", "strand")
  ))

dmr.sim <- ldply(dmr.sim, data.frame)

write.table(
  x = dmr.sim, file = gzfile("./output/methpipe_dmr_sim_data.txt.gz"),
  sep = "\t", quote = F, row.names = F
)
```

```{r}
datatable(
  dmr.sim,
  rownames = F,
  filter = "top", extensions = c("Buttons", "ColReorder"), options = list(
    pageLength = 10,
    buttons = c("copy", "csv", "excel", "pdf", "print"),
    colReorder = list(realtime = FALSE),
    dom = "fltBip"
  )
)
```


# SessionInfo

```{r}
devtools::session_info()
```