---
title: "pWGBSSimla: Defiant DMRs"
author: "Deepak Tanwar"
date: "<b>Created on:</b> 2020-02-14 <br> <b>Updated on:</b> `r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: tango
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    keep_md: no
    number_sections: no
    fig_width: 8
    fig_height: 8
    fig_caption: true
    df_print: paged
    code_folding: show
  fontsize: 12pt
  geometry: margin=1in
  documentclass: article
# bibliography: references.bib
link-citations: yes
---

# Packages required
```{r, warning=F, message=FALSE}
library(DT)
library(dplyr)
```

# Method description
Defiant computes p-values using weighted Welch's t-test, with weights giving larger importance to replicates with higher coverage. The initially found region is then widened, CpG by CpG, as long as the added CpG do not violate any condition (e.g. methylation difference high enough, coverage, p-value, etc.). If a stretch of CpGs that do not pass the criteria is met, the expansion is stopped.


# DMRs analysis

## Simulated data
```{r, engine='bash'}
mkdir -p ./output/sim_data && cd ./output/sim_data

DEFIANT=/home/ubuntu/softwares/downloads/defiant/defiant

$DEFIANT -c 1 -f -v -CpN 10 -p 0.9 -i <(zcat ../../input/sim_rep1.bed.gz),<(zcat ../../input/sim_rep3.bed.gz),<(zcat ../../input/sim_rep5.bed.gz) <(zcat ../../input/sim_rep2.bed.gz),<(zcat ../../input/sim_rep4.bed.gz),<(zcat ../../input/sim_rep6.bed.gz) 2> log >> log

cd ../../
```


# DMR Tables
Identified DMRs with q-value not higher than 0.05.


## Simulated data
```{r}
dmr.sim <- read.delim("./output/sim_data/set1_vs_set2_c1_CpN10_d1_p0.90_P10_q.tsv",
  header = T, sep = "\t", stringsAsFactors = F, check.names = F
)

dmr.sim[,1] <- paste0("chr", dmr.sim[,1])

# rename the q-value column to be compatible with further analysis
colnames(dmr.sim)[10] <- "qval"

# filter only regions with mean difference at least 10 (%)
dmr.sim <- dmr.sim[abs(dmr.sim$Mean_Difference) >= 10,]

write.table(dmr.sim, file = gzfile("./output/defiant_dmr_sim_data.txt.gz"), sep = "\t", quote = F, row.names = F)
```


```{r}
c2 <- dmr.sim$qval[dmr.sim$qval <= 0.05]
datatable(
  dmr.sim,
  rownames = F,
  filter = "top", extensions = c("Buttons", "ColReorder"), options = list(
    pageLength = 10,
    buttons = c("copy", "csv", "excel", "pdf", "print"),
    colReorder = list(realtime = FALSE),
    dom = "fltBip"
  )
) %>%
  formatStyle(
    "qval",
    backgroundColor = styleEqual(c2, rep("#66C2A5", length(c2)))
  )
```


# SessionInfo

```{r}
devtools::session_info()
```