---
title: "Filer Negative controls & generate simulated data"
author: "Deepak Tanwar & Hana Parizkova"
date: "<b>Created on:</b> 2018-11-25 <br> <b>Updated on:</b> `r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    highlight: pyg
    keep_md: no
    number_sections: yes
    fig_width: 8.5
    fig_height: 11
    fig_caption: true
    df_print: paged
    code_folding: show
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  fontsize: 12pt
  geometry: margin=1in
  documentclass: article
---

# Packages required
```{r 20181125-filtered-simulated-data-dmrseq-1, message=F, warning=F}
library(dmrseq)
library(data.table)
library(tidyverse)
library(dplyr)
```


# Negative control
```{r 20181125-filtered-simulated-data-dmrseq-2 }
# Files from 6 replicates
files <- list.files(path = "./input", pattern = "rep", full.names = T)
anno <- read.table("./input/phenoData.txt.gz", header = T, sep = "\t", stringsAsFactors = F)

# Read data
read.files <- lapply(files, function(x) {
  a <- fread(
    input = x, sep = "\t", data.table = F, header = F, stringsAsFactors = F,
    col.names = c("chr", "pos", "M", "cov")
  )
  colnames(a)[3:4] <- paste(strsplit(x, "\\.|/")[[1]][4], colnames(a)[3:4], sep = ".")
  return(a)
})

# Make one dataframe from list of files
tab <- Reduce(function(x, y) {
  merge(x, y, by = c("chr", "pos"))
}, read.files)
```


# BSseq object

## BSseq object creation and filtering
```{r 20181125-filtered-simulated-data-dmrseq-3 }
# Generate BSseq object
bs <- BSseq(
  chr = tab$chr, pos = tab$pos,
  M = data.matrix(tab[, grep(pattern = ".M", x = colnames(tab))]),
  Cov = data.matrix(tab[, grep(pattern = ".cov", x = colnames(tab))]),
  pData = anno, sampleNames = anno$names
)


# Keeping loci with with coverage at least 5 in at least two samples in each group
loci.idx <- which(rowSums(getCoverage(bs[, pData(bs)$Group == "Group1"], type = "Cov") >= 5) >= 2 &
  rowSums(getCoverage(bs[, pData(bs)$Group == "Group2"], type = "Cov") >= 5) >= 2)

bs.nc <- bs[loci.idx, ]

save(bs.nc, file = "./output/bsseq_neg_control.RData", compress = T)
```


## Saving filtered data
```{r 20181125-filtered-simulated-data-dmrseq-4 }
# Extract information from the object
chr.nc <- as.character(seqnames(bs.nc))
pos.nc <- start(bs.nc)
cov.nc <- data.frame(getCoverage(bs.nc, type = "Cov"))
M.nc <- data.frame(getCoverage(bs.nc, type = "M"))

for (i in 1:ncol(cov.nc)) {
  n <- paste0("./output/NC_rep", i, ".bed")
  df <- data.frame(chr.nc, pos.nc, M.nc[, i], cov.nc[, i])
  fwrite(x = df, file = n, sep = "\t", row.names = F, col.names = F, quote = F)
}
```



# Simulated data
- 100 simulated DMRs added to negative control data
- simulated data created by `simDMRs` function from `dmrseq` package

```{r 20181125-filtered-simulated-data-dmrseq-5 }
# reorder samples to create a null comparison
bs.null <- bs[, c(1, 3, 5, 2, 4, 6)]


# Set seeds to maintain reproducibility
set.seed(100)

# add 100 DMRs
bs.sim <- simDMRs(bs = bs.null, num.dmrs = 100)
bs.sim$bs <- bs.sim$bs[, c(1, 4, 2, 5, 3, 6)]

colnames(bs.sim$bs) <- str_replace(colnames(bs.nc) , "NC", "sim")
save(bs.sim, file = "./output/bsseq_sim_100.RData", compress = T)
```

## Simulated DMRs
```{r 20181125-filtered-simulated-data-dmrseq-6 }
dmr.ranges <- data.frame(bs.sim$gr.dmrs, stringsAsFactors = F)
sim.dmr <- data.frame(dmr.ranges, mean.cov = bs.sim$dmr.mncov, effect.size = bs.sim$delta, stringsAsFactors = F)
write.table(sim.dmr,
  file = gzfile("./output/simulated_DMRS.txt.gz", compression = 3),
  sep = "\t", quote = FALSE, row.names = FALSE
)
```


## Saving data in same format as of Negatie Control
```{r 20181125-filtered-simulated-data-dmrseq-7 }
# Re-arrange the data
sim <- bs.sim$bs

# Extract information from the object
chr.sim <- as.character(seqnames(sim))
pos.sim <- start(sim)
cov.sim <- data.frame(getCoverage(sim, type = "Cov"))
M.sim <- data.frame(getCoverage(sim, type = "M"))

for (i in 1:ncol(cov.sim)) {
  n <- paste0("./output/sim_rep", i, ".bed")
  df <- data.frame(chr.sim, pos.sim, M.sim[, i], cov.sim[, i])
  fwrite(x = df, file = n, sep = "\t", row.names = F, col.names = F, quote = F)
}
```


# phenoData tables

## Negative control
```{r 20181125-filtered-simulated-data-dmrseq-8 }
anno.nc <- anno[, 3:4]

write.table(anno.nc, "./output/anno_neg_control.txt", sep = "\t", quote = F)
```

## Simulated DMRs
```{r 20181125-filtered-simulated-data-dmrseq-9 }
anno.sim <- anno.nc %>%
  mutate_all(funs(str_replace(., "NC", "sim")))
rownames(anno.sim) <- anno.sim$names

write.table(anno.sim, "./output/anno_sim_data.txt", sep = "\t", quote = F)
```


# Compress all text outputs in output
```{r 20181125-filtered-simulated-data-dmrseq-10 }
system("pigz -11 -p 16 ./output/*.txt")
system("pigz -11 -p 16 ./output/*.bed")
```


# SessionInfo

<details>
```{r 20181125-filtered-simulated-data-dmrseq-11 }
devtools::session_info()
```
</details>
