---
title: "DMRs: BSseq"
author: "Hana Parizkova & Deepak Tanwar"
date: "<b>Created on:</b> 2018-11-25 <br> <b>Updated on:</b> `r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    highlight: pyg
    keep_md: no
    number_sections: yes
    fig_width: 8.5
    fig_height: 11
    fig_caption: true
    df_print: paged
    code_folding: show
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  fontsize: 12pt
  geometry: margin=1in
  documentclass: article
---

Analysis following the [Analyzing WGBS data with bsseq](https://bioconductor.org/packages/release/bioc/vignettes/bsseq/inst/doc/bsseq_analysis.html) vignette.

# Packages required
```{r, warning=F, message=FALSE}
library(bsseq)
library(DT)
```

# Method description
In BSseq, the genome is partitioned into windows and the methylation proportions in each window are approximated by a second degree polynomial. To find DMRs, signal-to-noise statistc on the smoothed data, similar to t-test, is computed.




# Input files and phenoData
Read in input files.
## Negative control
```{r}
pheno.nc <- read.table("./input/anno_neg_control.txt.gz", header = T, sep = "\t", stringsAsFactors = F)

files.nc <- list.files("./input", pattern = "NC_rep", full.names = T)
bs.nc <- read.bismark(
  files = files.nc, 
  colData = pheno.nc, 
  strandCollapse = F,
  nThread = detectCores() - 1
)

pData(bs.nc) <- pheno.nc
```

## Simulated data
```{r}
pheno.sim <- read.table("./input/anno_sim_data.txt.gz", header = T, sep = "\t", stringsAsFactors = F)

files.sim <- list.files("./input", pattern = "sim_rep", full.names = T)
bs.sim <- read.bismark(
  files = files.sim, sampleNames = pheno.sim$names, strandCollapse = F,
  mc.cores = detectCores() - 1
)

pData(bs.sim) <- pheno.sim
```


# Smoothing
The first step of the analysis is smoothing of the data.
```{r}
bs.smoothed.nc <- BSmooth(BSseq = bs.nc, mc.cores = detectCores() - 1)
bs.smoothed.sim <- BSmooth(BSseq = bs.sim, mc.cores = detectCores() - 1)
```


# Computing t-statistics
In the vignette, CpGs with no or little coverage are removed at this point. We have already filtered the data in the Data download and filtering step.

We now compute t-statistics.
```{r}
bs.tstat.nc <- BSmooth.tstat(bs.smoothed.nc,
  mc.cores = detectCores() - 1, estimate.var = "same",
  group1 = pheno.nc$names[pheno.nc$Group == "Group1"],
  group2 = pheno.nc$names[pheno.nc$Group == "Group2"],
  local.correct = TRUE,
  verbose = TRUE
)

bs.tstat.sim <- BSmooth.tstat(bs.smoothed.sim,
  mc.cores = detectCores() - 1, estimate.var = "same",
  group1 = pheno.sim$names[pheno.sim$Group == "Group1"],
  group2 = pheno.sim$names[pheno.sim$Group == "Group2"],
  local.correct = TRUE,
  verbose = TRUE
)
```


## Marginal distribution of the `t-statistic`
We can have a look at the marginal distribution of the t-statistics.
```{r, fig.align='center', fig.width=11, fig.height=8.5, fig.cap="Distribution of `t-statistic`."}
par(mfrow = c(1, 2), pty = "s")
plot(bs.tstat.nc)
title("Negative control")

plot(bs.tstat.sim)
title("Simulated data")
```


# DMRs
We can now compute DMRs by thresholding the t-statistics. Default values are used. The resulting DMRs are filtered for only those that are at least 10 CpGs long and the mean difference between the two groups is at least 0.1.
```{r}
# Negative control
dmr.nc <- dmrFinder(bs.tstat.nc)
dmr.nc <- subset(dmr.nc, n >= 10 & abs(meanDiff) >= 0.1)
write.table(dmr.nc,
  file = gzfile("./output/bsseq_dmr_neg_control.txt.gz"), sep = "\t",
  quote = F, row.names = F
)

# Simulated data
dmr.sim <- dmrFinder(bs.tstat.sim)
dmr.sim <- subset(dmr.sim, n >= 10 & abs(meanDiff) >= 0.1)
write.table(dmr.sim,
  file = gzfile("./output/bsseq_dmr_sim_data.txt.gz"), sep = "\t",
  quote = F, row.names = F
)
```

# DMR Tables

## Negative control
```{r}
datatable(
  dmr.nc,
  rownames = F,
  filter = "top", extensions = c("Buttons", "ColReorder"), options = list(
    pageLength = 10,
    buttons = c("copy", "csv", "excel", "pdf", "print"),
    colReorder = list(realtime = FALSE),
    dom = "fltBip"
  )
)
```

## Simulated data
```{r}
datatable(
  dmr.sim,
  rownames = F,
  filter = "top", extensions = c("Buttons", "ColReorder"), options = list(
    pageLength = 10,
    buttons = c("copy", "csv", "excel", "pdf", "print"),
    colReorder = list(realtime = FALSE),
    dom = "fltBip"
  )
)
```


# SessionInfo

<details>
```{r}
devtools::session_info()
```
</details>
